Recursive partitioning of linear growth curve models

### 2023-01-20 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
improvements in estimation section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3244 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 7b8f56f862b0187b7e2e45826d7934fc2a724b76

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the terminal nodes. Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \citep{FokkySmit18}.

All other methods for partitioning LGCMs take a fully local estimation approach: Within every terminal node, a full parametric model is estimated based on the observations in that node only. This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity. GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model. Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Therefore, the identity function can be used for the link $g$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

The parameters of a traditional GLMM can be estimated by, for example, (restricted) maximum likelihood. The GLMM tree model is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:

\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
	\item Set $r = r+1$. Estimate the partition using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with terminal node $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}

The algorithm initializes by assuming zero random effects. In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2. Thus, the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset. The fixed effects are estimated \textit{locally}, using the observations in the current node. Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges if the tree does not change from one iteration to the next.

>>>>

The main characteristic that sets GLMM trees apart from other methods for partitioning LGCMs is its local-global estimation approach: GLMM trees do not fit a full parametric model in each of the subgroups defined by the terminal nodes of the tree. Instead, the fixed-effects parameters are estimated \textit{locally}, using the observations within a terminal node, and the random-effects parameters are estimated \textit{globally}, using all observations. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits (i.e., intercepts only) in the terminal nodes. With GLMM trees, the approach was generalized to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of a wide range of possible interaction effects \citep{FokkySmit18}.

All other methods for partitioning LGCMs take a fully local estimation approach: Within every subgroup defined by the terminal nodes, a full parametric model is estimated based on the observations in that node only. This fully local estimation approach provides more flexibility but also comes at a higher computational burden and increased model complexity. GLMM trees require estimation of a (much) lower number of random-effects parameters, which likely reduces overfitting and improves stability and generalizability of the fitted model. Furthermore, the fully local estimation approach requires possible partitioning variables to be measured at the highest level of the multilevel structure, while the local-global estimation approach allows for partitioning variables to be measured at any level.

@@

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
Throughout this paper we focus on the case with a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. Therefore, the identity function can be used for the link $g$. Generalizations to other response variable types within the GLM are straightforward. Further, $X_i$ is the $n_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We further assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column of observed timepoints for subject $i$. Note that the number and spacing of observed timepoints may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on the subgroup/node~$j$ which observation~$i$ falls into. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$, and $b_i$ is a vector of random effects for subject $i$. Finally, $b$ is assumed to follow a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

The parameters of a traditional GLMM can be estimated, among other techniques, by maximum likelihood (ML) or restricted ML (REML).

Thus, when it is known which node~$j$ each observation~$i$ falls into, the GLMM specified by Equations~\ref{eq:expected_value}--\ref{eq:GLMMtree} can be fitted ``as usual'', yielding local subgroup-specific fixed-effect estimates $\hat \beta_j$ and global random-effect estimates $\hat b_i$. To infer the subgroup membership for all observations~$i$, the random-effect estimate is treated as a known offset and a GLM tree is estimated using the model-based (MOB) recursive partitioning algorithm of \cite{ZeilyHoth08}. The overall GLMM tree model is then estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm where the initialization simply assumes zero random effects.

\begin{enumerate}
\setlength\itemsep{0em}
	\setcounter{enumi}{-1}
	\item Initialize by setting $r$ and all values $\hat{b}_{(r)}$ to 0.
	\item Set $r = r+1$. Fit a GLM tree using $Z_{i} \hat{b_i}_{(r-1)}$ as an offset.
	\item Fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta_{j} + Z_{i} b_i$ with the partition $j(r)$ from Step~1. Extract posterior predictions $\hat{b}_{(r)}$ from the estimated model.
	\item Repeat Steps~1 and~2 until convergence.
\end{enumerate}

%% The algorithm initializes by assuming zero random effects. In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2. Thus, the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset. The fixed effects are estimated \textit{locally}, using the observations in the current node.

Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges when the partition~$j(r)$ from the GLM tree is the same as $j(r-1)$.
